A full approximation storage multilevel method \\ for nonlinear variational inequalities

### 2023-06-30 paper/fascd.tex
Start of some edits

\newcommand{\ip}[2]{\left<#1,#2\right>}

@@

@@

\begin{abstract}
The \emph{full approximation storage constraint decomposition} (FASCD) multilevel finite element method introduced here is a common extension of the full approximation storage multigrid technique for nonlinear partial differential equations, due to A.~Brandt, and of the constraint decomposition method introduced by X.-C.~Tai for the variational inequalities associated to inequality-constrained optimization.  We extend the constraint decomposition idea by recognizing and exploiting the telescoping nature of certain function space subset decompositions arising from multilevel mesh hierarchies.  Then, by using strong smoothers with work proportional to the number of unknowns on the finite element mesh level, we demonstrate that FASCD V-cycles show superior mesh-independent convergence rates and that FASCD F-cycles (full multigrid) are optimal solvers.  The several example problems include unilateral and bi-obstacle variational inequality problems with symmetric linear, nonsymmetric linear, and nonlinear differential operators.
\end{abstract}

@@

The constraint decomposition (CD) methods of Tai \cite{Tai2003} are designed to solve coercive variational inequality (VI) problems which arise from minimization of a convex functional over a convex set.  These methods have been shown to converge for multilevel and domain decompositions in a finite element (FE) context.  In the multilevel case, the CD method has been shown to have almost grid-independent error bounds, and near-optimal complexity, for elliptic, linear obstacle problems \cite[Subsection 5.4]{Tai2003}; see also \cite[Theorem 4.6 and Algorithm 4.7]{GraeserKornhuber2009}.  In this paper we extend the multilevel CD method by replacing the coarse-level corrections with a full approximation storage (FAS; \cite{Brandt1977,Bruneetal2015}) approach, and we apply this new solver to nonlinear VI problems which do not arise from minimization.

Specifically, our \emph{full approximation storage constraint decomposition} (FASCD) method extends the algorithms in \cite{GraeserKornhuber2009} and \cite{Tai2003} in the following directions:
\renewcommand{\labelenumi}{\emph{(\roman{enumi})}}
\begin{enumerate}
\item The FAS coarse corrections are adapted for general VI problems.  The needed computations require neither an objective function nor linearity of the residual operator.
\item Box constraints, namely upper and lower obstacles, are permitted; see Section \ref{sec:femultilevel}.
\item We observe an asymmetry special to multilevel CD methods, namely that up-smoothing is more effective than down-smoothing because a ``telescoping'' subset property can be exploited on the up-smoothing side.
\item We implement V-cycles (Algorithm \ref{alg:fascd}) and F-cycles (full multigrid; Algorithm \ref{alg:fascd-fcycle})
\item By using a Newton-type active set method as a smoother, with a fixed number of Krylov iterations per step so that the work of the smoother is proportional to the number of unknowns on the given level, we demonstrate that F-cycle FASCD is an optimal solver.
\end{enumerate}

FASCD V-cycles exhibit essentially the same mesh-independent convergence rates for VI problems that FAS multigrid V-cycles exhibit for the corresponding (unconstrained) PDE problems.  The F-cycle results in Section \ref{sec:results} show nearly-textbook multigrid efficiency, namely solutions within discretization error in work comparable to few smoother sweeps on the finest-level mesh \cite{BrandtLivne2011}, for classical Laplacian obstacle problems, $p$-Laplacian obstacle problems, and nonnegativity-constrained advection-diffusion equations.  Furthermore, these results are essentially insensitive to the length or structure of the free boundary, i.e.~the geometric complexity of the coincidence set.

The iterates from FASCD are admissible, and thus the nonlinear operator in the VI needs only be defined for admissible states.  For example, we solve an ice sheet geometry problem (Examples \ref{ex:sia}, \ref{ex:results:sia}) in which the ice flow formulas are only meaningful for surface elevation iterates which exceed the bed elevation.  (More generally, fluid-layer problems, with a non-negative thickness constraint, often require the solution of an auxiliary PDE on a domain determined inside the VI residual evaluation \cite{Bueler2021conservation}.  However, though a target for future applications of FASCD, such problems are not addressed here.)  Note that non-admissible methods, including semi-smooth methods \cite{BensonMunson2006}, would require unnatural modifications of the operator formula in such a fluid-layer context.  FASCD solves this low-regularity ice sheet problem, wherein the solution has unbounded gradient at the free boundary, by a mesh-independent number of F-cycle iterations.

This paper is organized as follows.  In Section \ref{sec:vi} we recall the theory of coercive and monotone variational inequalities, with several example problems.  Section \ref{sec:cd} reviews the basics of CDs, as proposed by Tai \cite{Tai2003}, including the associated iterations.  Section \ref{sec:femultilevel} sets up multilevel FE hierarchies, and then Section \ref{sec:cdmultilevel} presents defect constraints generated via monotone restriction operators \cite{GraeserKornhuber2009}.  Here we propose an apparently new understanding of telescoping sets for up-smoothing, based on an ``incomplete'' CD.  Section \ref{sec:vcycle} then states the FASCD V-cycle, with Section \ref{sec:implementation} addressing convergence criteria, F-cycles, and our smoother choice.  Section \ref{sec:results} gives FASCD numerical results from a parallel Firedrake \cite{Rathgeberetal2016} implementation for the problems introduced in Section \ref{sec:vi}.  Discussion and outlook in Section \ref{sec:discussion}, and an Appendix on how FASCD reduces to classical FAS multigrid for PDEs, finish the work.

@@

Suppose $\cV$ is a real and reflexive Banach space with norm $\|\cdot\|$ and topological dual space $\cV'$.  Denote the dual pairing of $\phi \in \cV'$ and $v\in\cV$ by $\ip{\phi}{v} = \phi(v)$, and define the (Banach space) norm on $\cV'$ by $\|\phi\|_{\cV'} = \sup_{\|v\|=1} |\ip{\phi}{v}|$.  Let $\cK \subset \cV$ be a nonempty, closed, and convex subset, the \emph{constraint set}; elements of $\cK$ are said to be \emph{admissible}.  For a continuous, but generally nonlinear, operator $f:\cK \to \cV'$ and a linear \emph{source functional} $\ell\in \cV'$ we consider the following \emph{variational inequality} (VI) problem, with solution $u^*\in \cK$ if it exists:
\begin{equation}
\ip{f(u^*)}{v-u^*} \ge \ip{\ell}{v-u^*} \qquad \text{for all } v\in \cK. \label{eq:vi}
\end{equation}
Because $f$ is (generally) nonlinear, the source term $\ell$ is not strictly needed in stating this class of problems.  In fact, by redefining $f$ we may take $\ell=0$.  However, allowing $\ell$ clarifies the algorithm in Section \ref{sec:vcycle}.

>>>>

\newcommand{\ip}[2]{\langle#1,#2\rangle}

@@

\usepackage{todonotes}
\newcommand{\pef}[1]{\todo[inline, color=red!20]{PF: #1}}

@@

\begin{abstract}
We present the \emph{full approximation storage constraint decomposition} (FASCD) multilevel method for solving variational inequalities. FASCD is a common extension of both the full approximation storage (FAS) multigrid technique for nonlinear partial differential equations, due to A.~Brandt, and of the constraint decomposition (CD) method introduced by X.-C.~Tai for variational inequalities arising in inequality-constrained optimization.  We extend the constraint decomposition idea by recognizing and exploiting the telescoping nature of certain function space subset decompositions arising from multilevel mesh hierarchies.  Then, by using strong smoothers with work proportional to the number of unknowns on the finite element mesh level, we demonstrate that FASCD V-cycles show excellent mesh-independent convergence rates and that FASCD F-cycles (full multigrid) are optimal solvers. The example problems include unilateral and bi-obstacle variational inequality problems with symmetric linear, nonsymmetric linear, and nonlinear differential operators; the examples demonstrate very good convergence even on problems where other multilevel VI solvers struggle.
\pef{I want to say it's better than monotone multigrid. Is this too subtle/too much?}

\end{abstract}

@@

The constraint decomposition (CD) methods of Tai \cite{Tai2003} are designed to solve coercive variational inequality (VI) problems arising as the Karush--Kuhn--Tucker optimality conditions for the minimization of a convex functional over a convex set.  These methods have been shown to converge for multilevel and domain decompositions in a finite element (FE) context; in the multilevel case, the CD method has been shown to have almost grid-independent error bounds, and near-optimal complexity, for elliptic, linear obstacle problems \cite[Subsection 5.4]{Tai2003}; see also \cite[Theorem 4.6 and Algorithm 4.7]{GraeserKornhuber2009}.  In this paper we extend the multilevel CD method by replacing the coarser-level corrections with a full approximation storage (FAS; \cite{Brandt1977,Bruneetal2015}) approach, and we apply this new solver to nonlinear VI problems which do not arise from minimization.
\pef{Note I've changed ``coarse-level corrections'' to ``coarser-level corrections'', to imply we're not just operating only on the coarsest grids, but also on intermediate grids.}

Specifically, our \emph{full approximation storage constraint decomposition} (FASCD) method extends the algorithms in \cite{GraeserKornhuber2009} and \cite{Tai2003} in the following directions:
\renewcommand{\labelenumi}{\emph{(\roman{enumi})}}
\begin{enumerate}
\item The FAS coarse corrections are adapted for general VI problems. The needed computations require neither an objective function nor linearity of the residual operator.
\item Box constraints, namely upper and lower obstacles, are permitted; see Section \ref{sec:femultilevel}.
\item We observe an asymmetry special to multilevel CD methods, namely that up-smoothing is more effective than down-smoothing because a ``telescoping'' subset property can be exploited on the up-smoothing side.
\item We implement V-cycles (Algorithm \ref{alg:fascd}) and F-cycles (full multigrid; Algorithm \ref{alg:fascd-fcycle}).
\item By using a Newton-type active set method as a smoother, with a fixed number of Krylov iterations per step so that the work of the smoother is proportional to the number of unknowns on the given level, we demonstrate that F-cycle FASCD is an optimal solver.
\pef{Clarify/limit which classes of problems we're claiming this for.}
\end{enumerate}

\pef{I think we need some text here hitting the nail on the head about what's different to monotone multigrid methods (even if it's just the absence of the invasive modifications of the basis functions).}

In our experiments, FASCD V-cycles exhibit essentially the same mesh-independent convergence rates for VI problems that FAS multigrid V-cycles exhibit for the corresponding (unconstrained) PDE problems.  The F-cycle results in Section \ref{sec:results} show nearly-textbook multigrid efficiency, namely solutions within discretization error in work comparable to few smoother sweeps on the finest-level mesh \cite{BrandtLivne2011}, for classical Laplacian obstacle problems, $p$-Laplacian obstacle problems, and nonnegativity-constrained advection-diffusion equations.  Furthermore, these results are essentially insensitive to the length or structure of the free boundary, i.e.~the geometric complexity of the coincidence set.

The iterates from FASCD are admissible, and thus the nonlinear operator in the VI needs only be defined for admissible states.  For example, we solve an ice sheet geometry problem (Examples \ref{ex:sia}, \ref{ex:results:sia}) in which the ice flow formulas are only meaningful for surface elevation iterates that do not penetrate the bedrock\footnote{More generally, fluid-layer problems, with a non-negative thickness constraint, often require the solution of an auxiliary PDE on a domain determined inside the VI residual evaluation \cite{Bueler2021conservation}.}.  Note that non-admissible methods, including semi-smooth methods \cite{BensonMunson2006}, would require unnatural and invasive modifications of the operator formula in such a fluid-layer context.  FASCD solves this low-regularity ice sheet problem, wherein the solution has unbounded gradient at the free boundary, with a mesh-independent number of F-cycle iterations.

This paper is organized as follows.  In Section \ref{sec:vi} we recall the theory of coercive and monotone variational inequalities, with several example problems.  Section \ref{sec:cd} reviews the basics of CDs, as proposed by Tai \cite{Tai2003}, and their induced iterations.  Section \ref{sec:femultilevel} sets up multilevel FE hierarchies, and then Section \ref{sec:cdmultilevel} presents defect constraints generated via monotone restriction operators \cite{GraeserKornhuber2009}.  Here we propose an apparently new understanding of telescoping sets for up-smoothing, based on an ``incomplete'' CD.  Section \ref{sec:vcycle} then states the FASCD V-cycle, with Section \ref{sec:implementation} addressing convergence criteria, F-cycles, and our smoother choice.  Section \ref{sec:results} gives FASCD numerical results from a parallel Firedrake \cite{Rathgeberetal2016} implementation for the problems introduced in Section \ref{sec:vi}.  The work concludes with discussion and outlook in Section \ref{sec:discussion}, and an appendix on how FASCD reduces to classical FAS multigrid for PDEs.

@@

Let $\cV$ be a real and reflexive Banach space with norm $\|\cdot\|$ and topological dual space $\cV'$.  Denote the dual pairing of $\phi \in \cV'$ and $v\in\cV$ by $\ip{\phi}{v} = \phi(v)$, and define the (Banach space) norm on $\cV'$ by $\|\phi\|_{\cV'} = \sup_{\|v\|=1} |\ip{\phi}{v}|$.  Let $\cK \subset \cV$ be a nonempty, closed, and convex subset, the \emph{constraint set}; elements of $\cK$ are said to be \emph{admissible}.  For a continuous, but generally nonlinear, operator $f:\cK \to \cV'$ and a linear \emph{source functional} $\ell\in \cV'$ we consider the following \emph{variational inequality} (VI) problem: find $u^*\in \cK$ such that
\begin{equation}
\ip{f(u^*)}{v-u^*} \ge \ip{\ell}{v-u^*} \quad \text{for all } v\in \cK. \label{eq:vi}
\end{equation}
Because $f$ is (generally) nonlinear, the source term $\ell$ is not strictly needed in stating this class of problems.  In fact, by redefining $f$ we may take $\ell=0$.  However, introducing $\ell$ clarifies the algorithm in Section \ref{sec:vcycle}.
